\documentclass[12pt]{article}
\usepackage{techart}

\parindent=0pt
\parskip=8pt

\begin{document}
\title{STAT 651 Midterm 1}
\author{}
\maketitle

\textbf{Name}: \underline{\ \ \hskip8cm\ \ }
\medskip

Answer the following questions.  Be sure to show your work. 
Closed book, closed notes.  You may use two 8 $\times$ 11 piece of
paper containing notes you have written.
Calculators are allowed.
Time limit: 60 minutes.
\medskip

\begin{enumerate}

\item(15 pts)  Consider a hand of poker, in which 5 cards are selected at random from a deck of 52 cards.
  \begin{enumerate}
  \item Find the number of hands that make a full house (3 of a kind and 2 of a kind).
  \item Find the number of hands that make a flush (5 cards of the same suit).
  \item Which type of hand is less likely to occur (and so should have higher value
    in designing a game)?
  \end{enumerate}

\item(20 pts)  Consider an experiment in which one 4-sided die and one fair coin are tossed.
  \begin{enumerate}
  \item Describe the sample space, $S$; how many elements does $S$ have?
%  \item How many items does $\cal{B}$, the usual $\sigma$-field have?
  \item Let $X$ be a random variable associated with the experiment, where:
%  If $\omega \in S$, we define
  $$X = \left\{ \begin{array}{ll}
  j & {\mbox{ if coin is H and we see }} j {\mbox{ spots}} \\
  2j & {\mbox{ if coin is T and we see }} j {\mbox{ spots}} \\
  \end{array}\right.$$
    \begin{enumerate}
    \item Find $f_X(x)$, the pmf for $X$.
    \item Sketch the graph of $F_X(x)$, the cdf for $X$.
    \end{enumerate}
  \end{enumerate}
  
\item (10 pts)  Let $A$ and $B$ be events that are independent {\underline{and}} mutually exclusive.
\begin{enumerate}
\item Show that either $P(A)=0$ or $P(B)=0.$
\item Is it necessarily true that either $A=\emptyset$ or $B=\emptyset$?
    Explain briefly, or describe a counter example.
\end{enumerate}

\item (15 pts) It is known that 95\% of children like the movie, 
Curse of the Were Rabbit (CoTWR), i.e.\ P(enjoy  movie $|$ child) = .95.
But (just) 90\% of adults like it.  The general population consists of
40\% children, 60\% adults.
If a randomly selected person from the population
likes CoTWR, what is the probability that they are a child?

\item (25 pts) Let $X$ be a random variable that has cdf $F_X(x)$, given by:
$$F_X(x) = \left\{ \begin{array}{ll}
 1-\exp(-3x) & {\mbox{ if }} x \ge 0 \\
 0 & {\mbox{ if }} x < 0.
 \end{array} \right. $$
  \begin{enumerate}
  \item Find $f_X(x)$, the pdf of $X$.
  \item Let $Y = e^X$.
    \begin{enumerate}
    \item Find $f_Y(y)$, the pdf of $Y$. (What is the range of values that $Y$ can take?)
    \item Show that the pdf of $Y$ integrates to 1.
    \item Find $E\{Y\}$.
    \end{enumerate}
  \end{enumerate}
 
\item (15 pts) Let $X \sim {\mbox{Bernoulli}}(p)$, where $0 \le p \le 1.$
  (I.e.\, $P(X=1)=p$ and $P(X=0)=1-p.$)
  \begin{enumerate}
  \item Find $E\{X^m\}$, where  $m = 1,2,3,\ldots.$ (Hint: First, find the pmf of $X$.)
  \item Find $E\{e^{tX}\}$, where $t$ is a known constant, $-\infty < t < \infty$.
  \item Use the result above to find
    $E\{e^{-X}\}$ and $E\{e^{3X}\}$.
  \end{enumerate}


\end{enumerate}


\end{document}

\item The variance of an RV $X$ is defined as var($X$) = $E[(X-\mu_X)^2]$, where $\mu_X=E[X]$.
Let $a$ and $c$ be constants, $-\infty < a < \infty$, $-\infty < c < \infty$.
  \begin{enumerate}
  \item Show that var($aX$) = $a^2 {\mbox{var}}(X)$.
  \item Show that var($X+c$) = var($X$).
  \end{enumerate}
