\documentclass[12pt]{article}
\usepackage{techart}
\begin{document}

\title{STAT 651 HW 1}
\author{Due Friday, 9/17}
\date{}
\maketitle

\begin{enumerate}
\item
For each of the following experiments, describe the sample space.
Also state whether the sample space is finite, countably infinite, or uncountably infinite.
\begin{enumerate}
\item Toss a coin four times.
\item Count the number of insect-damaged leaves on a plant.
\item Measure the lifetime (in hours) of a particular brand of light bulb.
\item Record the weights of 10-day-old rats.
\item Observe the proportion of defectives in a shipment of electronic components.
\end{enumerate}

\item
For events $A$ and $B$, find formulas for the probabilities of the following
events in terms of the quantities $P(A)$, $P(B)$ and $P(A\cap B)$.
\begin{enumerate}
\item either $A$ or $B$ or both.
\item either $A$ or $B$ but not both.
\item at least one of $A$ or $B$.
\item at most one of $A$ or $B$.
\end{enumerate}

\item
Approximately 1/3 of all human twins are identical (one-egg) and
2/3 are fraternal (two-egg) twins.  Identical twins are necessarily the same sex,
with male and female being equally likely.  Among fraternal twins, approximately 1/4 are
both female, 1/4 are both male, and half are one male and one female.  Finally,
among all U.S.\ births, approximately 1 in 90 is a twin birth.  Define the following
events:
\[
\begin{array}{rcl}
A & = & {\mbox{\{a U.S.\ birth results in twin females\}}} \\
B & = & {\mbox{\{a U.S.\ birth results in identical twins\}}} \\
C & = & {\mbox{\{a U.S.\ birth results in twins\}}} \\
\end{array}
\]
\begin{enumerate}
\item State, in words, the event $A \cap B \cap C$.
\item Find $P(A\cap B \cap C)$.
\end{enumerate}

\item
A phone company offers six wireless plans, each with a different number
of Anytime minutes. An experiment consists of selecting a customer and
recording the plan number and whether the customer was under or over the
number of Anytime minutes last month.
\begin{enumerate}
\item Find the sample space $\mathcal{S}$ for this experiment.
\item Consider the following events:\\
$A$: the customer was over the number of Anytime minutes.\\
$B$: The customer is in plan 1, 2, or 3.\\
$C$: the customer is in plan 5 or 6, and was over the number of Anytime
minutes.\\
$D$: the customer is in an even-numbered plan.\\
Find the outcomes in each of the following events:\\
(i) $B^c$ \ \  (ii) $A \cup B$ \ \  
(iii) $C \cap D$ \ \  (iv) $(A \cap D)^c$
\end{enumerate}

\item
Consider the type of clothes dryer (gas or electric) purchased by each
of five different customers at a certain store.
\begin{enumerate}
\item If the probability that at most one of these customers
purchases an electric dryer is .428, what is the probability that at
least two purchase an electric dryer?
\item If $P(\text{all five purchase gas}) = .116$ and
$P(\text{all five purchase electric}) = .005$, what is the probability
that at least one of each type is purchased?
\end{enumerate}

\item
An art gallery has 20 stored paintings, but has just made room to
display 7 of them. Seven paintings will be randomly selected and offered
to the public for sale.
\begin{enumerate}
\item How many different collections of 7 paintings are possible?
\item Suppose 10 of the 20 stored works are by the same local artist. What is
the probability all seven of the selected paintings will be by this
artist?
\end{enumerate}


\item
Find (calculate) the number of elements there are in the sigma algebra $\cal{B}$ for
the following sample spaces $S$:
\begin{enumerate}
\item $S$ = \{ outcomes obtained by flipping 4 distinct coins once \} (e.g. penny, nickel, dime, quarter)
\item $S$ = \{ outcomes obtained by flipping 4 identical coins once \} (e.g. four 1919 pennies)
\item $S$ = \{ outcomes obtained by flipping 4 coins once, where two of the coins are (say)
    indistinguishable pennies, and the other two are (say) indistinguishable quarters \}
\item $S$ = \{ outcomes obtained by flipping a single coint 4 times \}
\item $S$ = \{ outcomes obtained by flipping $N = n_1 + n_2$ coins once, where $n_1$ are of one type, indistinguishable
from one another, and $n_2$ are of another type, indistinguishable from one another \}.
\end{enumerate}

\end{enumerate}


\end{document}

