\documentclass[12pt]{article}
\usepackage{techart}

\begin{document}
\title{STAT 651 HW 10}
\author{Due Friday, 12/10}
\date{}
\maketitle


\begin{enumerate}
\item Let $X$ and $Y$ be independent uniform(0,1) RV's. Find
$P(X/Y\le t)$ and $P(XY\le t)$. (Pictures will help.)

\item Let $(X,Y)$ follow a multivariate normal distribution with parameters
$\mu_X = 5$, $\mu_Y = -1$, $\sigma_X = 1$, $\sigma_Y=4$ and $\rho$ (to be specified momentarily).
(You can use the matrix formulas learned in class.)

  \begin{enumerate}
  \item Find the conditional distribution of $Y$ given $X=7$, assuming $\rho = 0.6$.
  Is the conditional variance of $Y$ greater than or less than the unconditional (marginal) variance?
  \item Find the conditional distribution of $Y$ given $X=7$, assuming $\rho = -0.6$.
  Is the conditional variance of $Y$ greater than or less than the unconditional (marginal) variance?
  \end{enumerate}

\item Let $X$, $Y$ and $Z$ be independent Poisson random variables with parameters
$\lambda_X$, $\lambda_Y$ and $\lambda_Z$ respectively. Let $U = X+Z$ and $V = Y+Z$.
Find $\cov(U,V)$ and $\operatorname{cor}(U,V)$.

\item Let $\vec{X} \sim \text{Multinomial}(7; .4,.3,.2,.1)$.  For
example, $\vec{X}$ represents
the number of red, green, blue, and white balls drawn when drawing samples of size 7
with replacement from an urn containing
4 red, 3 green, 2 blue, and 1 white balls.  Let $f(\vec{x})$ be the corresponding pmf.
  \begin{enumerate}
  \item Find $f(3,2,1,1)$.
  \item Find $f(1,1,4,1)$. (So more blue balls than there are in the urn.)
  \item Find $f(0,3,2,2)$.
  \item (Optional) For how many vectors of $\vec{x}$ is $f(\vec{x})$ non-zero?
  \end{enumerate}

\item Jensen's Inequality states that if $g(x)$ is convex on the support of $X$,
 then $E(g(X)) \ge g(E(X)).$
  \begin{enumerate}
  \item Show that if $X \sim N(-1,\sigma^2)$, then $E(|X|)\ge 1.$
  What bound would you get if $X \sim N(-10,\sigma^2)$?  What if $X \sim N(3,\sigma^2)$?
  \item Show that if $X \sim $Poisson($\lambda$) then $E(X^3) \ge \lambda^3.$ (Can you find the
exact value of $E(X^3)$?)
  \item Use Jensen's Inequality to show that if $X \sim N(0,\sigma^2)$,
   then $E(X^4) \ge (\sigma^2)^2 = \sigma^4.$
  \end{enumerate}

\item Let $X_i \stackrel{iid}{\sim} $ Exp($\beta$), $i=1,2,\ldots,n$.
Find the distribution of $X_{(1)}$, the first order statistic, and
of $X_{(n)}$, the $n$-th order statistic.  Do either of these correspond to a familiar distribution?

\item
Let $X_i \stackrel{iid}{\sim} N(\mu_X,\sigma^2)$, $i=1,2,\ldots,n$ and
$Y_j \stackrel{iid}{\sim} N(\mu_Y,\sigma^2)$, $j = 1,\ldots,m$.
Assume that the $\{X_i\}$ and $\{Y_j\}$ are independent.
     \begin{enumerate}
     \item Find the sampling distribution of $\bar{X}-\bar{Y}$.
     \item Find the distribution of $\frac{1}{\sigma^2}[(n-1)S_X^2 + (m-1)S_Y^2]$.
     \end{enumerate}


\item Let $X_i \stackrel{iid}{\sim} N(0,1)$, $i=1,\ldots,5.$
Let $X_6$ be another independent observation, also $N(0,1)$.
Let $\bar{X} = (X_1 + X_2 + \cdots + X_5)/5$, i.e.\ the mean of just the first 5 $X$'s.
 What is the distribution of
\begin{enumerate}
\item $W = \sum_1^5 X_i^2$? Why?
\item $U = \sum_1^5 (X_i-\bar{X})^2$? Why?
\item $\sum_1^5 (X_i-\bar{X})^2+X_6^2$?  Why?
\item $\sqrt{5}X_6/\sqrt{W}$? Why?
\item $2X_6/\sqrt{U}$? Why?
\item $2(5\bar{X}^2 + X_6^2)/U$? Why?
\end{enumerate}

\end{enumerate}

\end{document}
