\documentclass[12pt]{article}
\usepackage{techart}

\begin{document}
\title{STAT 651 HW 2}
\author{Due Friday, 9/20}
\date{}
\maketitle

\begin{enumerate}
\item
Counting the number of various types of hands in poker
(drawing 5 cards randomly, without replacement).
Count the number of ways that a hand can be drawn consisting of
    \begin{enumerate}
    \item 3 of a kind + 2 of a kind (full house; ``kind'' means the
        face number)
%    \item 3 of a kind plus two singletons (i.e. non-pair)
    \item two pairs
    \item five singletons (no pairs, no triplets, etc.)
    \end{enumerate}

    (Hint: answers are 3744, %54912,
     123552, and 1317888)

\item
Two players, $A$ and $B$, alternately and independently flip a coin
and the first player to obtain a head wins.
Assume player $A$ flips first.
  \begin{enumerate}
  \item If the coin is fair, what is the probability that A wins?
  \item If $P(\text{head}) = p$, not necessarily $1/2$,
    what is the probability that $A$ wins?
  \item Show that for all $p \in (0, 1)$, $P(\text{$A$ wins}) > 1/2$.
  \end{enumerate}

  (Hint: Try to write $P(\text{$A$ wins})$
  in terms of the events $E_1, E_2,\dotsc$,
  where $E_i$ is the event that
  $A$ wins at her $i$th toss.)

\item
Suppose that 5\% of men and .25\% of women are color-blind.
A person is chosen at random and that person is color-blind.
What is the probability that the person is male?
(Assume males and females to be in equal numbers.)

\item
If the probability of hitting a target is $1/5$, and 10 shots are fired
independently.
    \begin{enumerate}
    \item
    What is the probability of the target being hit at least twice?
    \item
    What is the conditional probability that the target is hit at least twice,
    given that it is hit at least once?
    \end{enumerate}

\item
Prove each of the following statements.
(Assume that any conditioning event has positive probability.)
  \begin{enumerate}
  \item If $P(B) = 1$ then $P(A\given B) = P(A)$ for any event $A$.
  \item If $A \subset B$ then $P(B\given A) = 1$ and
    $P(A\given B)=P(A) / P(B)$.
  \item If $A$ and $B$ are mutually exclusive, then
    $P(A\given A\cup B) = P(A) \bigm/ \bigl(P(A) + P(B)\bigr)$.
  \end{enumerate}

\item
A pair of events $A$ and $B$ cannot be simultaneously
\emph{mutually exclusive} and \emph{independent}.
Prove that if $P(A) > 0$ and $P(B) > 0$, then:
  \begin{enumerate}
  \item If $A$ and $B$ are mutually exclusive, they cannot be independent.
  \item If $A$ and $B$ are independent, they cannot be mutually exclusive.
  \end{enumerate}

\item
If $S = \{ \text{sequences of 4 consecutive tosses of a single coin} \}$.
Find the cdf and pmf for the random variable $X$, where
$X(\omega) = (\text{\# heads}) \times (\text{\# tails})$
for $\omega \in S$.
\end{enumerate}

\end{document}
